---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data
```{r openfile, echo=TRUE}
library(tidyverse)
library(xtable)
activ_monit_dat <- read_csv("./activity/activity.csv", na = "NA")
```
```{r no scientific notation,echo=FALSE}
options(scipen = 999)
```

## What is mean total number of steps taken per day?
```{r steps complete dataset, echo=TRUE}
steps_per_day <- activ_monit_dat %>% 
  group_by(date) %>% 
  summarise(steps_day = sum(steps,na.rm = T))
```

#### *Total Steps per day:*

```{r, echo=TRUE}
steps_per_day
```


### Histogram of total number of steps per day

```{r,echo=TRUE}
p1 <- ggplot(steps_per_day,aes(steps_day)) +
  geom_histogram(color = "blue") +
  labs(title = "Total Number of Steps Per Day",
       subtitle = "with missing data") +
  xlab("number of steps")
  
p1
```

#### *Median of total number of steps per day*

```{r mediansteps,echo=TRUE}
med_steps <- median(steps_per_day$steps_day)
```

**Median** of total number of steps per day is **`r med_steps`**

#### *Mean of total number of steps per day*

```{r meansteps, echo=TRUE}
aver_steps <- mean(steps_per_day$steps_day)
```

**Mean** of total number of steps per day is **`r aver_steps`**

## What is the average daily activity pattern?

```{r aver steps across all days,echo=TRUE}
aver_daily_pattern <- activ_monit_dat %>% 
  group_by(interval) %>% 
  summarise(average_steps_per_interval = mean(steps,na.rm=T))
  
ggplot(aver_daily_pattern ,aes(x=interval,y=average_steps_per_interval))+
  geom_line() +
  labs(title = "Average daily Active Pattern") +
  xlab("5 minute interval") + 
  ylab("Average number of steps taken")
```
```{r max pattern of steps,echo=TRUE}
max_pat_n_steps <- aver_daily_pattern %>% 
  filter(average_steps_per_interval == max(average_steps_per_interval)) %>% select(interval)
```
The 5-minute interval which contains the maximum number is `r max_pat_n_steps`

## Imputing missing values

```{r missing,echo=TRUE}
missing = sum(!complete.cases(activ_monit_dat))
```

Total number of **missing values** in the dataset is **`r missing`**

*replace na values in steps variable by the mean of total steps*
*and store this new complete dataset in activ_complete dataframe*

```{r replace na steps,echo=TRUE}
activ_complete <- activ_monit_dat %>%
  mutate(steps = replace_na(steps,mean(steps,na.rm = TRUE)))
```

*Check if there are any missing values left in steps variable, as well as in interval variable*

```{r check for missing,echo=TRUE}
sum(is.na(activ_complete$steps))
sum(is.na(activ_complete$interval))
```

*Aggregate complete dataset to daily*

```{r complete dataset to daily, echo=TRUE}
steps_complete <- activ_complete %>% 
  group_by(date) %>% 
  summarise(steps_day = sum(steps,na.rm = T))
```

*mean and median calculation for complete dataset*

```{r mean and median complete dataset,echo=TRUE}

average_steps_complete <- round(mean(steps_complete$steps_day),1)
med_steps_complete <- round(median(steps_complete$steps_day),1)
```

Complete dataset **mean** value is **`r average_steps_complete`** and **median** **`r med_steps_complete`**. It can be seen that median and mean have the same value, denoting that by replace missing values with the mean, distribution approximates normal distribution, compared to the raw dataset.

*Data distribution of the total number of steps taken each day on a histogram using complete dataset*

```{r histogram of complete dataset,echo=TRUE}
p2 <- ggplot(steps_complete,aes(steps_day)) +
  geom_histogram(color = "red") +
  labs(title = "Total Number of Steps Per Day",
       subtitle = "complete dataset") +
  xlab("number of steps")

p2
```

*comparison of steps taken per day distributions (dataset with missing values, complete dataset). Histograms created with ggplot as well as cowplot packages*

```{r steps distribution comparison,echo=TRUE}
library(cowplot)

plot_grid(p1,p2, ncol = 2)

```



## Are there differences in activity patterns between weekdays and weekends?

wday function from **lubridate** package is used, for dividing dates to weekdays and weekends. *Monday* corresponds to *1* and Sunday, as the last day of the week, to *7*. For identifying activity patterns between the levels, complete dataset is used. 

```{r patterns weekdays/weekends,echo=TRUE}
library(lubridate)

activity_complete_steps <- activ_complete %>% 
  mutate(dayofweek = wday(date),
         weekends = ifelse(dayofweek>=6,"weekend","weekday"),
         weekends = as.factor(weekends))

head(steps_complete)
```

Using **lattice** package we create a panel plot containing a time series plot  of the 5-minute interval and the average number of steps taken, averaged across all weekday days or weekend days. 

```{r panel plot,echo=TRUE}

activity_complete_per_interval <- activity_complete_steps %>% 
  group_by(interval,weekends) %>% 
  summarise(steps_5min = mean(steps))

library(lattice)

xyplot(steps_5min~interval | weekends,data=activity_complete_per_interval,
       type="l",
       layout = c(1,2), ylab = "average steps",
       xlab = "5 minutes interval",
       main = "Time Series plot - Average number of Steps on weekends & weekdays")

```
